Measurement of a dwarf galaxy rotational velocity from its Lyman-alpha line emission

### 2016-09-01 paper/letter.tex
BH is back

\newcommand{\tol}{Tololo 1214-277}
\newcommand{\lya}{Ly$\alpha$}
\newcommand{\hb}{H$\beta$}
\newcommand{\ha}{H$\alpha$}
\newcommand{\oiii}{[OIII]}
\newcommand{\oii}{[OII]}
\newcommand{\nii}{[NII]}
\newcommand{\esca}{erg cm$^{-2}$ s$^{-1}$ \AA$^{-1}$}
\newcommand{\esc}{erg cm$^{-2}$ s$^{-1}$}
\newcommand{\es}{erg s$^{-1}$}
\newcommand{\esa}{erg s$^{-1}$}
\newcommand{\kms}{km s$^{-1}$}
\newcommand{\apj}{ApJ}  
\newcommand{\jcap}{JCAP}  
\newcommand{\apjs}{ApJS}  
\newcommand{\apjl}{ApJL}  
\newcommand{\aj}{AJ}  
\newcommand{\mnras}{MNRAS}  
\newcommand{\mnrassub}{MNRAS accepted}  
\newcommand{\aap}{A\&A}  
\newcommand{\aaps}{A\&AS}  
\newcommand{\araa}{ARA\&A}  
\newcommand{\nat}{Nature}  
\newcommand{\physrep}{PhR}
\newcommand{\pasp}{PASP}    
\newcommand{\pasj}{PASJ}    
\newcommand{\sigmaclump}{$69^{+17}_{-15}$ km s$^{-1}$}
\newcommand{\inftyclump}{$79^{+73}_{-42}$ km s$^{-1}$}
\newcommand{\probaclump}{$0.78^{+0.16}_{-0.43}$}
\def\simgt{\lower.5ex\hbox{\gtsima}}
\def\simlt{\lower.5ex\hbox{\ltsima}}
%------------------------------------------------------

\begin{document}
\pagestyle{empty}
\noindent
\textbf{Lyman-$\alpha$ emission reveals a dispersion-supported dwarf}
\\
\\
Jaime E. Forero-Romero$^{1}$, Max Groenke$^2$, Mar\'ia Camila
Remolina-Guti\'errez$^1$, Juan Nicol\'as
Garavito-Carmargo$^3$, Mark Dijkstra$^2$.
\\
\\
\scriptsize
{$^1$ Departamento de Física, Universidad de los Andes, Cra. 1
  No. 18A-10 Edificio Ip, CP 111711, Bogot\'a, Colombia 
\\
$^2$ Institute of Theoretical Astrophysics, University of Oslo,
Postboks 1029 Blindern, NO-0315 Oslo, Norway.
\\
$^3$ Department of Astronomy, University of Arizona, 933 North Cherry
Avenue, Tucson, AZ 85721, USA. 
\normalsize
\\
\\
\textbf{
  Star-forming Compact Dwarf Galaxies (CDGs) resemble the expected
  pristine conditions of the first galaxies in the Universe.    
Before the observational detection of the first galaxies becomes
reality, CDGs are the best systems to test our ideas on primordial
galaxy formation and evolution.    
Here we report on one of such CDGs, \tol, which presents
a broad symmetric Lyman-$\alpha$ emission that had evaded theoretical
interpretation so far. 
In this letter we explain these features by two different models: an
homogeneous gaseous sphere undergoing bulk rotation and an interstellar
medium composed by outflowing clumps with additional random motions.
It is the first time that an observed \lya\ spectrum can be explained
assuming either of these physical conditions.
We find that both models independently require high velocities
(either a bulk rotation of $348^{+75}_{-48}$ km s$^{-1}$ or a clump velocity
dispersion of \sigmaclump with outflows of
$79^{+73}_{-42}$ km s$^{-1}$) consistent with a dynamical mass at
least $10$ times larger than its baryonic mass.  
We argue that the most plaussible explanation is provided by the
multiphase model as it is a close match to the galaxy scaling relationship
for a dark matter dominated dwarf elliptical supported by its
velocity dispersion. 
This work demonstrates the importance of considering multiphase
physics among the possible conditions shaping the \lya\ spectra of the
first galaxies.   Additionally, if future kinematic maps of
\tol\ confirm the high velocities postulated in our model, it would
help to stablish the \lya\ line morphology as an observational tracer to
study the co-evolution of the first generation of galaxies and its
host dark matter halos. 
}

@@

However, the physical interpretation of Ly-$\alpha$ observations is
not straightforward. 
This is due to the resonant nature of the Ly-$\alpha$ line. 
A Ly-$\alpha$ photon follows a diffusion-like process before escaping
the galaxy or being absorbed by dust. 
The resulting line profile becomes sensitive to the dynamical, chemical
and thermal conditions in the interstellar medium. 
There are very few analytically tools available to interpret the
Ly-$\alpha$ line.
They are applicable only in very few cases of highly symmetrical
conditions, which are hardly met in real astrophysical systems.
For these reasons the interpretation of Ly-$\alpha$ observations
require state-of-the-art Monte Carlo radiative transfer simulations.

\tol\ is a compact star forming dwarf galaxy that presents a
strong Ly-$\alpha$ emission \cite{Thuan97} with two puzzling 
features: it is symmetric and single peaked.
Commonly, the Ly-$\alpha$ line has a single or asymmetric double peak. 
These two special features in \tol\ cannot be explained with
conventional models.

In this letter we show how the \tol's \lya\ profile can be explained
either by rotation or the recently developed class of more complex
multiphase models. These models predict a wider variety of spectra
including, single, double and triply peaked spectra.  
Figure 1. summarizes our findings.
Dots represent the observational data for \tol\ with the
overplot from our best fit models from the analytical solution for a
rotating homogeneous gas sphere (continuous line) and a clumpy ISM
(dashed line).
This is the first time that these models have been introduced with
success to explain the a \lya\ profile.

@@

In the multiphase model the best constrained parameters are
the clump velocity dispersion  $\sigma_{\rm{cl}}=$\sigmaclump\
the clumps outflowing velocity $v_{\infty, \rm{cl}}=$\inftyclump\
and the fraction of the \lya\ emission that is  coming
from the cold clumps  $P_{\rm cl}=$\probaclump.
The multiphase model assumes that the clouds are distributed over a
sphere of $5$kpc in radius, close to the $\approx 4$kpc physical size
of \tol\ as determined by optical imaging.
The assumed physical size and the velocity dispersion $\sigma_{\rm
  cl}$ correspond to a  dynamical mass of  $M_{\rm
  dyn}=2.8^{+1.3}_{-1.2}\times 10^{9}$ $M_{\odot}$. In the rotation
model assuming reasonable bounds for the number density of neutral Hydrogen atoms (
$0.01<n/\mathrm{atoms\ cm^{-3}}<0.1$) and using $\tau=10^7$ the radius
of the emission region can be bracketed to be in $0.55 <
r_s/\mathrm{kpc}< 5.5$.
Using $V_{\rm max}=300$km s$^{-1}$, we find a dynamical mass range of
$5.45\times 10^{9}M_{\odot} < M_{\rm dyn} < 5.45\times 10^{10} M_{\odot}$.

Which solution is more plausible? If \tol\ followed the fundamental
plane relationship between its mean surface brightness $I_e$, the
projected half-light radius $R_e$ and the velocity dispersion
$\sigma$, described by $\log I_e=1.6 \log\sigma - 1.21\log R_e +
0.55$ \cite{2009ApJ...698.1590G,2011ApJ...726..108T},  
the expected velocity dispersion would be on the order of $47\pm 3$
\kms, which only presents a small tension with the multiphase model.   
The solution of a high rotational velocity pushes \tol\ outside the
scaling relationship for dwarfs. One could account for the high
velocities if the galaxy had a supermassive black hole accounting for
the mass that cannot be explained by baryonic or dark mass. The economy of
assumptions makes it more reasonable to expect that the multiphase
model could be a better match to reality.

A future observational test to clarify the physical nature of
\tol\ would require integral field unit measurements spatially
resolving its spatial extent. 
\tol\ spans a region of $4$ arcseconds,
an instrument such as the Multi Unit Spectroscopic Explorer with its
nominal $0.2$ arcseconds spatial sampling over a $1.0$ arcminute field
in wide-field mode could provide a coarse mapping of different
ionization lines to infer a kinematic map.
Another observational test includes the measurement of the escape
fraction of \lya\ continuum ionizing radiation. 
In the rotational model this fraction should be zero, while
the multiphase model predicts that in this case it should be around
$0.5^{+1.0}_{-0.4}$\%.

All in all, the mere existence of a strong LAE galaxy with a broad,
symmetric line is interesting.
It raises the question whether some high redshift LAEs have asymmetric
lines because the blue half was truncated by the intergalactic medium.
In this case the \lya\ radiation could emerge as a low surface
brightness glow, which may be connected to \lya\ halos, while also
influencing the way LAEs can be used as a probe of reionization.

These findings demonstrate the importance of including multiphase
conditions as a feature to model the \lya\ line in high redshift
galaxies.
If the estimated velocity dispersion proves to be consistent with
future observational kinematic maps, it could provide a new way to
study velocity dispersions in distant gas rich starbursting galaxies, 
opening the possibility to study the dynamical co-evolution of
galaxies and dark matter halos in the first generation of galaxies.

>>>>

\newcommand{\tol}{Tololo 1214-277}
\newcommand{\lya}{Ly$\alpha$}
\newcommand{\hb}{H$\beta$}
\newcommand{\ha}{H$\alpha$}
\newcommand{\oiii}{[OIII]}
\newcommand{\oii}{[OII]}
\newcommand{\nii}{[NII]}
\newcommand{\esca}{erg cm$^{-2}$ s$^{-1}$ \AA$^{-1}$}
\newcommand{\esc}{erg cm$^{-2}$ s$^{-1}$}
\newcommand{\es}{erg s$^{-1}$}
\newcommand{\esa}{erg s$^{-1}$}
\newcommand{\kms}{km s$^{-1}$}
\newcommand{\apj}{ApJ}  
\newcommand{\jcap}{JCAP}  
\newcommand{\apjs}{ApJS}  
\newcommand{\pasa}{PASA}  
\newcommand{\apjl}{ApJL}  
\newcommand{\aj}{AJ}  
\newcommand{\mnras}{MNRAS}  
\newcommand{\mnrassub}{MNRAS accepted}  
\newcommand{\aap}{A\&A}  
\newcommand{\aaps}{A\&AS}  
\newcommand{\araa}{ARA\&A}  
\newcommand{\nat}{Nature}  
\newcommand{\physrep}{PhR}
\newcommand{\pasp}{PASP}    
\newcommand{\pasj}{PASJ}    
\newcommand{\sigmaclump}{$69^{+17}_{-15}$ km s$^{-1}$}
\newcommand{\inftyclump}{$79^{+73}_{-42}$ km s$^{-1}$}
\newcommand{\probaclump}{$0.78^{+0.16}_{-0.43}$}
\def\simgt{\lower.5ex\hbox{\gtsima}}
\def\simlt{\lower.5ex\hbox{\ltsima}}
%------------------------------------------------------

\begin{document}
\pagestyle{empty}
\noindent
\textbf{Lyman-$\alpha$ emission sketches a massive dwarf}
\\
\\
Jaime E. Forero-Romero$^{1}$, Max Groenke$^2$, Mar\'ia Camila
Remolina-Guti\'errez$^1$, Juan Nicol\'as
Garavito-Carmargo$^3$, Mark Dijkstra$^2$.
\\
\\
\scriptsize
{$^1$ Departamento de Física, Universidad de los Andes, Cra. 1
  No. 18A-10 Edificio Ip, CP 111711, Bogot\'a, Colombia 
\\
$^2$ Institute of Theoretical Astrophysics, University of Oslo,
Postboks 1029 Blindern, NO-0315 Oslo, Norway.
\\
$^3$ Department of Astronomy, University of Arizona, 933 North Cherry
Avenue, Tucson, AZ 85721, USA. 
\normalsize
\\
\\
\textbf{
  Star-forming Compact Dwarf Galaxies (CDGs) resemble the expected
  pristine conditions of the first galaxies in the Universe.    
Before the observational detection of the first galaxies becomes
reality, CDGs are the best systems to test our ideas on primordial
galaxy formation and evolution.    
Here we report on one of such CDGs, \tol, which presents
a broad symmetric Lyman-$\alpha$ emission that had evaded theoretical
interpretation so far. 
In this letter we explain these features by two different models: an
homogeneous gaseous sphere undergoing bulk rotation and an interstellar
medium composed by outflowing clumps with additional random motions.
It is the first time that an observed \lya\ spectrum can be explained
assuming either of these physical conditions.
We find that both models independently require high velocities
(either a bulk rotation of $348^{+75}_{-48}$ km s$^{-1}$ or a clump velocity
dispersion of \sigmaclump with outflows of
$79^{+73}_{-42}$ km s$^{-1}$) consistent with a dynamical mass at
least $10$ times larger than its baryonic mass.  
We argue that the most plausible explanation for this excess of
dynamical mass is the presence of a supermassive black hole at the
center of \tol. 
This work demonstrates the importance of considering multiphase
physics and rotation among the possible conditions shaping the
\lya\ spectra of the first galaxies.  
Additionally, if future kinematic maps of \tol\ confirm the high
velocities postulated in our model, it would provide new
evidence for dwarf galaxies as hosts of supermassive black
holes.  
}

@@

However, the physical interpretation of Ly-$\alpha$ observations is
not straightforward \cite{2015ApJ...805...14R}. 
This is due to the resonant nature of the Ly-$\alpha$ line. 
A Ly-$\alpha$ photon follows a diffusion-like process before escaping
the galaxy or being absorbed by dust. 
The resulting line profile becomes sensitive to the dynamical, chemical
and thermal conditions in the interstellar medium. 
There are very few analytically tools available to interpret the
Ly-$\alpha$ line.
They are applicable only in very few cases of highly symmetrical
conditions, which are hardly met in real astrophysical systems.
For these reasons the interpretation of Ly-$\alpha$ observations
require state-of-the-art Monte Carlo radiative transfer simulations.

\tol\ is a compact star forming dwarf galaxy that presents a
strong Ly-$\alpha$ emission \cite{Thuan97} with two puzzling 
features: it is symmetric and single peaked.
Commonly, the Ly-$\alpha$ line has a single or asymmetric double peak. 
These two special features in \tol\ cannot be explained with
conventional models \cite{2006A&A...460..397V,2015ApJ...812..123G}.

In this letter we show how the \tol's \lya\ profile can be explained
either by rotation or the recently developed class of more complex
multiphase models that predict a wider variety of spectra
including, single, double and triply peaked spectra.  
Figure 1. summarizes our findings.
Dots represent the observational data for \tol\ with the
overplot from our best fits from the analytical solution for a
rotating homogeneous gas sphere (continuous line) and the multiphase
model (dashed line). 
This is the first time that these models have been introduced with
success to explain an observed \lya\ profile.

@@

In the multiphase model the best constrained parameters are
the clump velocity dispersion  $\sigma_{\rm{cl}}=$\sigmaclump\,
the clump's outflowing velocity $v_{\infty, \rm{cl}}=$\inftyclump\
and the fraction of the \lya\ emission that is  coming
from the cold clumps  $P_{\rm cl}=$\probaclump.
The multiphase model assumes that the clouds are distributed over a
sphere of $5$kpc in radius, close to the $\approx 4$kpc physical size
of \tol\ as determined by optical imaging.
The assumed physical size and the velocity dispersion $\sigma_{\rm
  cl}$ correspond to a  dynamical mass of  $M_{\rm
  dyn}=2.8^{+1.3}_{-1.2}\times 10^{9}$ $M_{\odot}$. In the rotation
model assuming reasonable bounds for the number density of neutral Hydrogen atoms (
$0.01<n/\mathrm{atoms\ cm^{-3}}<0.1$) and using $\tau=10^7$ the radius
of the emission region can be bracketed to be in $0.55 <
r_s/\mathrm{kpc}< 5.5$.
Using $V_{\rm max}=300$km s$^{-1}$, we find a dynamical mass range of
$5.45\times 10^{9}M_{\odot} < M_{\rm dyn} < 5.45\times 10^{10} M_{\odot}$.

Are these values atypical? If \tol\ followed the fundamental
plane relationship between its mean surface brightness
$I_e$, the projected half-light radius $R_e$ and the velocity dispersion
$\sigma$, described by $\log I_e=1.6 \log\sigma - 1.21\log R_e +
0.55$ \cite{2009ApJ...698.1590G},  the expected velocity dispersion
should be on the order of $5 \pm 1$ 
\kms, which is a factor of $\sim$ $12$ - $60$ lower than the 
results from the multiphase-rotation models, respectively. 
These are equivalente to factors of $\sim$ $140$ - $3600$ on the
dynamical mass;
\tol\ seems to be significantly more massive than the expectation from
scaling relationships for dwarf ellipticals.
A factor of $100$ in dynamical mass could be accounted by dark matter if
\tol\ was hosted by a dark matter halo of  $\sim 10^{16}$ M$_{\odot}$
\cite{2011ApJ...726..108T}, which is very unlikely as it would mean that
\tol\ is at the center of a super rich cluster of galaxies,
contradicting observational evidence.
The remaining possibility is that \tol\ hosts a supermassive black hole of
$\sim 5\times 10^{9}$ M$_{\odot}$, corresponding to $10$ times its
estimated stellar and gaseous mass.

A new observational test is needed to clarify the physical nature of
\tol. 
We suggest that integral field unit measurements spatially
resolving its spatial extent is up to the task. 
\tol\ spans a region of $4$ arcseconds,
an instrument such as the Multi Unit Spectroscopic Explorer
\cite{2014Msngr.157...13B} with its
nominal $0.2$ arcseconds spatial sampling over a $1.0$ arcminute field
in wide-field mode could provide a coarse mapping of different
ionization lines to infer a kinematic map.
Another observational test includes the measurement of the escape
fraction of \lya\ continuum ionizing radiation. 
In the rotational model this fraction should be zero, while
the multiphase model predicts that averaging over all sightlines
it should be around $0.5^{+1.0}_{-0.4}$\%, with the possibility of strong
variations depending on viewing angle.

All in all, the mere existence of a strong LAE galaxy with a broad,
symmetric line is interesting.
It raises the question whether some high redshift LAEs have asymmetric
lines because the blue half was truncated by the intergalactic medium.
In this case the \lya\ radiation could emerge as a low surface
brightness glow, which may be connected to \lya\ halos, while also
influencing the way LAEs can be used as a probe of reionization
\cite{2014PASA...31...40D}.

These findings demonstrate the importance of including rotation and multiphase
conditions as features to model the \lya\ line in high redshift
galaxies.
Additionally, if the hypothesis of a supermassive black
hole in \tol\ proves to be consistent with future observational
kinematic maps, it could correspond to a so far undetected black hole
in a dwarf galaxy, providing a new way to test and probe
theories on the co-evolution of galaxies and black holes in the first
generation of galaxies.
